Make manuscript figures

### 2022-06-11 scripts/R/make_manuscript_figures.rmd
add trees to example simulation figure

FIGDIR <- "figures"
OUTDIR <- "output"

@@

Load data
```{r}
accuracy_simulation_results <- read.delim("output/2021-08-31_estimator_accuracy_MLE.txt")
tpr_simulation_results <- read.delim("output/2021-08-31_GWAS_TPR_MLE.txt")
example_simulation_results <- read.delim("output_revisions/2022-05-30_simulateFullDataset_2.txt")
tree <- ape::read.tree(file = paste(OUTDIR, "pathogen_no_outgroup.newick", sep = "/"))
traits <- read.csv(file = paste(OUTDIR, "poumm_corrected_traits.csv", sep = "/"))
tree_w_trait_data <- treeio::full_join(tree, traits, by = "label")
POUMM_fit <- get(load(file = paste(OUTDIR, "POUMM_fit.RData", sep = "/")))
poumm_parameter_estimates <- read.csv(file = paste(OUTDIR, "poumm_parameter_estimates.csv", sep = "/"))
pc_scores <- read.csv(file = paste(OUTDIR, "shcs_vs_hapmap_pc_scores.csv", sep = "/"))
gwas_results <- read.delim(file = paste(OUTDIR, "gwas_results.pvals.effectsize.txt", sep = "/"), sep = "", header = T)
spvl_values <- read.csv(file = paste(OUTDIR, "spvl_values.csv", sep = "/"))
```

@@

top_row <- plot_grid(NULL, rmse_plot, rel_widths = c(1.03, 3), labels = c("", "A"))
png(filename = "figures/simulation_results.png", width = linewidth, height = height, unit = "cm", res = 300)
plot_grid(top_row, tpr_plot, align = "v", axis = "tb", nrow = 2, labels = c("", "B"))
dev.off()
```

@@

Example trait values from successful simulation
```{r}
p1 <- ggplot(data = example_simulation_results %>% rename("Heritability" = H2), aes(x = h, y = h.MWA)) +
      geom_point() +
      facet_grid(. ~ Heritability, labeller = "label_both") + 
      shared_theme +
      labs(x = "Simulated host part of trait", y = ehat_name)
ggsave(plot = p1, filename = paste(FIGDIR, "example_simulation_trait_values.png", sep = "/"),
      width = linewidth * 2/3, height = height * 2/3, units = "cm")
```

@@

>>>>

FIGDIR <- "figures"
OUTDIR <- "output"
OUTDIR_XANTHAMONAS <- "output_arabidopsis_xanthomonas"

@@

Load data
```{r}
accuracy_simulation_results <- read.delim("output_revisions/2022-06-11_estimator_accuracy_MLE.txt")
tpr_simulation_results <- read.delim("output_revisions/2022-06-11_GWAS_TPR_MLE.txt")
example_simulation_results <- read.delim("output_revisions/2022-06-11_simulateFullDataset.txt")
tree <- ape::read.tree(file = paste(OUTDIR, "pathogen_no_outgroup.newick", sep = "/"))
traits <- read.csv(file = paste(OUTDIR, "poumm_corrected_traits.csv", sep = "/"))
tree_w_trait_data <- treeio::full_join(tree, traits, by = "label")
POUMM_fit <- get(load(file = paste(OUTDIR, "POUMM_fit.RData", sep = "/")))
poumm_parameter_estimates <- read.csv(file = paste(OUTDIR, "poumm_parameter_estimates.csv", sep = "/"))
pc_scores <- read.csv(file = paste(OUTDIR, "shcs_vs_hapmap_pc_scores.csv", sep = "/"))
gwas_results <- read.delim(file = paste(OUTDIR, "gwas_results.pvals.effectsize.txt", sep = "/"), sep = "", header = T)
gwas_results_xanthomonas <- read.delim(file = paste(OUTDIR_XANTHAMONAS, "gwas_results.pvals.effectsize.txt", sep = "/"), sep = "", header = T)
spvl_values <- read.csv(file = paste(OUTDIR, "spvl_values.csv", sep = "/"))
```

@@

top_row <- plot_grid(NULL, rmse_plot, rel_widths = c(1.03, 3), labels = c("", "A"))
png(filename = "figures/simulation_results_v2.png", width = linewidth, height = height, unit = "cm", res = 300)
plot_grid(top_row, tpr_plot, align = "v", axis = "tb", nrow = 2, labels = c("", "B"))
dev.off()
```

@@

Example trait values from successful simulation
```{r}
# Re-generate simulation tree, attach trait data to it
set.seed(1)
simulation_tree <- simulateHIVTreeExpBL(500)
for (idx in 501:550) {
        single_clade <- ape::extract.clade(phy = simulation_tree, node = idx)
        if (length(single_clade$tip.label) > 8 && length(single_clade$tip.label) < 30) {
                print(idx)
        }
}
single_clade <- ape::extract.clade(phy = simulation_tree, node = 516)

example_data <- example_simulation_results %>% rename("label" = "sample", "Heritability" = "H2")
example_data_lo_h2 <- example_data %>% filter(Heritability == 0.15)
example_data_hi_h2 <- example_data %>% filter(Heritability == 0.65)

# Plot trait values on a clade in the tree
mar <- 20
lo_h2_tree <- ggtree(single_clade) %<+% example_data_lo_h2 + 
        geom_tippoint(aes(color = v), size = 3) + 
        scale_color_viridis_b(name = "Pathogen\npart of trait") + 
        theme_tree(plot.margin=margin(mar, mar, mar, mar))

hi_h2_tree <- ggtree(single_clade) %<+% example_data_hi_h2 + 
        geom_tippoint(aes(color = v), size = 3) + 
        scale_color_viridis_b(name = "Pathogen\npart of trait") + 
        theme_tree(plot.margin=margin(mar, mar, mar, mar))

# Plot trait value accuracy
p1 <- ggplot(data = example_data, aes(x = h, y = h.MWA)) +
      geom_point() +
      facet_grid(. ~ Heritability, labeller = "label_both") + 
      shared_theme +
      labs(x = "Simulated host part of trait", y = ehat_name)

# Arrange plots together
top_row <- plot_grid(
        lo_h2_tree + theme(legend.position = "none"),
        hi_h2_tree + theme(legend.position = "none"),
        labels = c("A", "B"),
        nrow = 1
)
plot_grid(top_row, get_legend(lo_h2_tree),
          p1,
          nrow = 2, align = "v", axis = "tb",
          rel_heights = c(1, 1.5),
          rel_widths = c(1, 0.2),
          labels = c("", "", "C"))
ggsave(filename = paste(FIGDIR, "example_simulation_trait_values.png", sep = "/"),
      width = linewidth, height = height * 1.3, unit = "cm")
```

@@

Q-Q plot of Arabidopsis-Xanthomonas GWAS results
```{r}
gwas_results_xanthomonas <- gwas_results_xanthomonas %>% filter(!is.na(P_standard))
qq_data_standard <- data.frame(
        observed_log10p = -log10(sort(gwas_results_xanthomonas$P_standard, decreasing = F)),
        expected_log10p = -log10(ppoints(nrow(gwas_results_xanthomonas))),
        gwas_type = standard_gwas_name)
qq_data_phylo <- data.frame(
        observed_log10p = -log10(sort(gwas_results_xanthomonas$P_corrected, decreasing = F)),
        expected_log10p = -log10(ppoints(nrow(gwas_results_xanthomonas))),
        gwas_type = corrected_gwas_name)
qq_results <- rbind(qq_data_standard, qq_data_phylo)

p <- ggplot(data = qq_results, aes(x = expected_log10p, y = observed_log10p)) +
        geom_point() +
        geom_abline(slope = 1, intercept = 0, linetype = "dashed",
                    color = RColorBrewer::brewer.pal(n = 3, name = diverging_palette)[[1]]) +
        facet_wrap(. ~ gwas_type) +
        labs(x = expression("expected log"[10]*"(p)"), y = expression("observed log"[10]*"(p)")) +
        shared_theme
ggsave(plot = p, filename = paste(FIGDIR, "qq_plots_xanthomonas.png", sep = "/"),
       width = linewidth, height = height * 2/3, units = "cm")
```

Manhattan plot of Arabidopsis-Xanthomonas GWAS results
```{r}
pt_size <- 1

# Add genome position to data
chr_lengths <- gwas_results_xanthomonas %>%
        group_by(CHROM) %>%
        summarise(chr_len = max(POS))
chr_starts <- chr_lengths %>%
        mutate(CHR_start = cumsum(as.numeric(chr_len)) - chr_len) %>%
        select(-chr_len)
gwas_results_annotated <- left_join(gwas_results_xanthomonas, chr_starts, by = "CHROM")
gwas_results_annotated <- gwas_results_annotated %>%
        arrange(CHROM, POS) %>%
        mutate(BP_position = CHR_start + POS)

# This is maybe not the best way, but makes sure tick marks and labels are at the center of each chromosome
x_axis_ticks <- chr_starts
max_BP_position <- max(gwas_results_annotated[gwas_results_annotated$CHROM == 5, "BP_position"])
x_axis_ticks <- rbind(
        x_axis_ticks,
        data.frame(
                CHROM = c(5, 6),
                CHR_start = c(max_BP_position, max_BP_position + 1000)))
x_axis_ticks$midpoint <- x_axis_ticks$CHR_start + (lead(x_axis_ticks$CHR_start) - x_axis_ticks$CHR_start)/2

# Make comparative GWAS overview plot
bonferroni_signifiance_threshold <- 0.05 / nrow(gwas_results_annotated)
overview_comparison <- ggplot(
        data = results_long %>% sample_frac(0.1),  # TODO: plot all of results, not sub-sample of them
        aes(x = BP_position, y = -log10(P))) +
        geom_point(aes(color = CHROM %% 2 == 0), size = pt_size) +
        scale_color_manual(values = c("grey", "dark grey")) +
        scale_x_continuous(
                breaks = x_axis_ticks$midpoint,
                labels = x_axis_ticks$CHROM,
                expand = c(0.01, 0.01)) +
        scale_y_continuous(limits = c(0, 16), expand = c(0, 0)) +
        labs(x = "Chromosome", y = expression('-Log'[10]*'(p)')) +
        geom_hline(
                yintercept = -log10(bonferroni_signifiance_threshold),
                linetype = "dashed") +
        shared_theme +
        theme(
                panel.grid.major.x = element_blank(),
                panel.grid.minor.x = element_blank(),
                legend.position = "none"
        ) +
        facet_grid(. ~ gwas_type)
ggsave(
        plot = overview_comparison,
        filename = paste(FIGDIR, "gwas_results_xanthomonas.png", sep = "/"),
        width = linewidth, height = height / 2, unit = "cm"
)
```

Get the most highly-associated and most changed SNPs from the Xanthomonas GWAS results
```{r}
# Most highly associated
most_highly_associated_standard <- gwas_results_annotated %>% top_n(n = 5, wt = -P_standard)
most_highly_associated_corrected <- gwas_results_annotated %>% top_n(n = 5, wt = -P_corrected)

most_highly_associated <- rbind(
        most_highly_associated_standard %>% mutate(type = "strongest association standard GWAS"),
        most_highly_associated_corrected %>% mutate(type = "strongest association corrected GWAS"))

tex_table_1 <- most_highly_associated %>%
        ungroup() %>%
        select(CHROM, POS, ID, BETA_standard, P_standard, BETA_corrected, P_corrected)

print(xtable(tex_table_1, type = "latex", display=c("s", "s", "s", "s", "g","g","g","g")),
      file = paste(FIGDIR, "strongest_associations_xanthamonas.tex", sep = "/"), math.style.exponents = TRUE)

write.csv(x = most_highly_associated_mhc_ccr5,
          file = paste(OUTDIR, "strongest_associations_xanthamonas.csv", sep = "/"), row.names = F)

# Most changed: higher significance
most_changed_positive <- gwas_results_annotated %>%
        group_by(CHROM) %>%
        top_n(n = 1, wt = -neg_log10_P_standard_minus_corrected)

write.csv(x = most_changed_positive,
          file = paste(OUTDIR, "most_increased_association_xanthomonas.csv", sep = "/"), row.names = F)
```
